Randomized trace estimation for parameter-dependent matrices applied to spectral density approximation

### 2025-02-17 paper/analysis.tex
adapt nomenclature and fix README

\paragraph{Nyström estimator.} Given a good low-rank approximation of a matrix, its trace can be cheaply estimated by simply taking the trace of that low-rank approximation. In the context of  spectral density estimation, the use of a non-negative smoothing kernel implies that $\mtx{B}(t)$ is symmetric positive definite for every $t$. This makes the Nyström method
from~\cite{gittens-2013-revisiting-nystrom} very well for constructing a (randomized) low-rank approximation.
Given a Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$, Nyström computes the approximation
\begin{equation}
    \mtx{B}(t) \approx \Nystr{\mtx{\Omega}}{\mtx{B}}(t) := (\mtx{B}(t) \mtx{\Omega}) (\mtx{\Omega}^{\top} \mtx{B}(t) \mtx{\Omega})^{\dagger} (\mtx{B}(t) \mtx{\Omega})^{\top}.
    \label{equ:nystrom-approximation}
\end{equation}
Related randomized low-rank approximations for general parameter-dependent matrices have been analyzed in~\cite{kressner-2024-randomized-lowrank}.
Exploiting the invariance of the trace under cyclic permutations, taking the trace of~\eqref{equ:nystrom-approximation} gives rise to the Nyström estimator
\begin{equation}
    \Trace(\mtx{B}(t)) \approx \Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t)) = \Trace( (\mtx{\Omega}^{\top} \mtx{B}(t) \mtx{\Omega})^{\dagger} ( \mtx{\Omega}^{\top} \mtx{B}(t)^2 \mtx{\Omega})).
    \label{equ:nystrom-trace-estimator}
\end{equation}
This estimator forms the basis of the so-called spectrum sweeping method \cite{lin-2017-randomized-estimation}.

@@

\begin{remark}
    A variant of \refthm{thm:hutchinson} can be derived when assuming that $\mtx{B}(t)$ is symmetric positive semi-definite and nonzero for every $t \in [a, b]$. After dividing both sides of \refequ{equ:moment-bound-hutchinson} by $\Trace(\mtx{B}(t))$, one can apply the remaining steps of the proof to deduce that
    \begin{equation*}
        \int_{a}^{b} \frac{| \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) |}{\Trace(\mtx{B}(t))} ~\mathrm{d}t < \varepsilon
    \end{equation*}
    holds with probability at least $1 - \delta$ if $n_{\mtx{\Psi}} = \mathcal{O}(\log(\delta^{-1}) \varepsilon^{-2} \rho^2)$. Here, 
    the quantity $
        \rho := \frac{1}{b-a} \int_{a}^{b} \frac{\lVert \mtx{B}(t) \rVert _F}{\Trace(\mtx{B}(t))} ~\mathrm{d}t \in [1,n]
    $
is small when the singular values of $\mtx{B}(t)$ decay slowly for the majority of $t\in [a,b]$ and large otherwise. This parallels an analogous result for constant matrices; see~\cite[Remark 2]{cortinovis-2022-randomized-trace}.
\end{remark}

@@

\begin{theorem}{Nyström estimator for parameter-dependent matrices}{nystrom}
    Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be symmetric positive semi-definite and continuous for $t \in [a, b]$. For integers $r \geq 2$ and $n_{\mtx{\Omega}} \geq r + 4$, consider the Nystr\"om approximation $\Nystr{\mtx{\Omega}}{\mtx{B}}(t)$ defined in~\refequ{equ:nystrom-approximation} with an $n\times n_{\mtx{\Omega}}$ Gaussian random matrix $\mtx{\Omega}$. Then, for any $\gamma \geq 1$, the bound 
    \[
        \int_{a}^{b} \big| \Trace(\mtx{B}(t)) - \Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t)) \big| ~\mathrm{d}t
        \leq \gamma^2 (1 + r) \int_{a}^{b} \sum_{i = r+1}^{n} \sigma_i(\mtx{B}(t)) ~\mathrm{d}t.
    \]
    holds with probability at least $1 - \gamma^{-(n_{\mtx{\Omega}} - r)}$.
\end{theorem}
\begin{proof}
Using that $\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)$ is positive semi-definite (see, e.g.,~\cite[Lemma 2.1]{frangella-2023-randomized-nystrom}) we obtain
    \begin{align*}
        \big| \Trace(\mtx{B}(t)) - \Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t)) \big|
        & = \big| \Trace(\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)) \big|
        = \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _{\ast} \\
        & = \lVert (\mtx{I}_n - \mtx{\Pi}_{\mtx{B}(t)^{\sfrac{1}{2}} \mtx{\Omega}}) \mtx{B}(t)^{\sfrac{1}{2}} \rVert _F^2,
    \end{align*}
    where $\lVert \cdot \rVert _{\ast}$ denotes the nuclear norm and we used \cite[Theorem 1]{gittens-2011-spectral-norm} in the last equality. This allows us to apply \cite[Theorem 5]{kressner-2024-randomized-lowrank}, which immediately implies the statement claimed by the theorem.
\end{proof}

@@

\begin{theorem}{Nyström++ trace estimator for parameter-dependent matrices}{nystrom-pp}
    Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be symmetric positive semi-definite and continuously depend on $t \in [a, b]$. If $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-1} \log(\delta^{-1}))$ with \emph{even} $n_{\mtx{\Omega}} \geq 4$, then for $\delta \in (0, 2e^{-1})$ and $\varepsilon \in (0, 1)$ with probability at least $1 - \delta$ 
    \begin{equation}
        \int_{a}^{b} | \Trace(\mtx{B}(t)) - \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B}(t)) |~\mathrm{d}t
        \leq \varepsilon \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t.
    \end{equation}
\end{theorem}

>>>>

\paragraph{Nyström estimator.} Given a good low-rank approximation of a matrix, its trace can be cheaply estimated by simply taking the trace of that low-rank approximation. In the context of  spectral density estimation, the use of a non-negative smoothing kernel implies that $\mtx{B}(t)$ is symmetric positive semi-definite (SPSD) for every $t$. This makes the Nyström method
from~\cite{gittens-2013-revisiting-nystrom} very well for constructing a (randomized) low-rank approximation.
Given a Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$, Nyström computes the approximation
\begin{equation}
    \mtx{B}(t) \approx \Nystr{\mtx{\Omega}}{\mtx{B}}(t) := (\mtx{B}(t) \mtx{\Omega}) (\mtx{\Omega}^{\top} \mtx{B}(t) \mtx{\Omega})^{\dagger} (\mtx{B}(t) \mtx{\Omega})^{\top}.
    \label{equ:nystrom-approximation}
\end{equation}
Related randomized low-rank approximations for general parameter-dependent matrices have been analyzed in~\cite{kressner-2024-randomized-lowrank}.
Exploiting the invariance of the trace under cyclic permutations, taking the trace of~\eqref{equ:nystrom-approximation} gives rise to the Nyström estimator
\begin{equation}
    \Trace(\mtx{B}(t)) \approx \Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t)) = \Trace( (\mtx{\Omega}^{\top} \mtx{B}(t) \mtx{\Omega})^{\dagger} ( \mtx{\Omega}^{\top} \mtx{B}(t)^2 \mtx{\Omega})).
    \label{equ:nystrom-trace-estimator}
\end{equation}
This estimator forms the basis of the so-called spectrum sweeping method \cite{lin-2017-randomized-estimation}.

@@

\begin{remark}
    A variant of \refthm{thm:hutchinson} can be derived when assuming that $\mtx{B}(t)$ isSPSD and nonzero for every $t \in [a, b]$. After dividing both sides of \refequ{equ:moment-bound-hutchinson} by $\Trace(\mtx{B}(t))$, one can apply the remaining steps of the proof to deduce that
    \begin{equation*}
        \int_{a}^{b} \frac{| \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) |}{\Trace(\mtx{B}(t))} ~\mathrm{d}t < \varepsilon
    \end{equation*}
    holds with probability at least $1 - \delta$ if $n_{\mtx{\Psi}} = \mathcal{O}(\log(\delta^{-1}) \varepsilon^{-2} \rho^2)$. Here, 
    the quantity $
        \rho := \frac{1}{b-a} \int_{a}^{b} \frac{\lVert \mtx{B}(t) \rVert _F}{\Trace(\mtx{B}(t))} ~\mathrm{d}t \in [1,n]
    $
is small when the singular values of $\mtx{B}(t)$ decay slowly for the majority of $t\in [a,b]$ and large otherwise. This parallels an analogous result for constant matrices; see~\cite[Remark 2]{cortinovis-2022-randomized-trace}.
\end{remark}

@@

\begin{theorem}{Nyström estimator for parameter-dependent matrices}{nystrom}
    Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be SPSD and continuous for $t \in [a, b]$. For integers $r \geq 2$ and $n_{\mtx{\Omega}} \geq r + 4$, consider the Nystr\"om approximation $\Nystr{\mtx{\Omega}}{\mtx{B}}(t)$ defined in~\refequ{equ:nystrom-approximation} with an $n\times n_{\mtx{\Omega}}$ Gaussian random matrix $\mtx{\Omega}$. Then, for any $\gamma \geq 1$, the bound 
    \[
        \int_{a}^{b} \big| \Trace(\mtx{B}(t)) - \Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t)) \big| ~\mathrm{d}t
        \leq \gamma^2 (1 + r) \int_{a}^{b} \sum_{i = r+1}^{n} \sigma_i(\mtx{B}(t)) ~\mathrm{d}t.
    \]
    holds with probability at least $1 - \gamma^{-(n_{\mtx{\Omega}} - r)}$.
\end{theorem}
\begin{proof}
Using that $\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)$ is SPSD (see, e.g.,~\cite[Lemma 2.1]{frangella-2023-randomized-nystrom}) we obtain
    \begin{align*}
        \big| \Trace(\mtx{B}(t)) - \Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t)) \big|
        & = \big| \Trace(\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)) \big|
        = \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _{\ast} \\
        & = \lVert (\mtx{I}_n - \mtx{\Pi}_{\mtx{B}(t)^{\sfrac{1}{2}} \mtx{\Omega}}) \mtx{B}(t)^{\sfrac{1}{2}} \rVert _F^2,
    \end{align*}
    where $\lVert \cdot \rVert _{\ast}$ denotes the nuclear norm and we used \cite[Theorem 1]{gittens-2011-spectral-norm} in the last equality. This allows us to apply \cite[Theorem 5]{kressner-2024-randomized-lowrank}, which immediately implies the statement claimed by the theorem.
\end{proof}

@@

\begin{theorem}{Nyström++ trace estimator for parameter-dependent matrices}{nystrom-pp}
    Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be SPSD and continuously depend on $t \in [a, b]$. If $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-1} \log(\delta^{-1}))$ with \emph{even} $n_{\mtx{\Omega}} \geq 4$, then for $\delta \in (0, 2e^{-1})$ and $\varepsilon \in (0, 1)$ with probability at least $1 - \delta$ 
    \begin{equation}
        \int_{a}^{b} | \Trace(\mtx{B}(t)) - \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B}(t)) |~\mathrm{d}t
        \leq \varepsilon \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t.
    \end{equation}
\end{theorem}

### 2025-02-17 paper/intro.tex
adapt nomenclature and fix README

All methods considered in this work are based on the following simple idea: Apply an existing randomized trace estimator to $\Trace(\mtx{B}(t))$ with \emph{constant} random vectors, that is, the same randomization is used for each value of the parameter $t$. \textcolor{green}{For example, the Girard-Hutchinson estimator becomes $n_{\mtx{\Psi}}^{-1} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{B}(t) \vct{\psi}_j$ for $n_{\mtx{\Psi}}$ constant standard Gaussian random vectors $\vct{\psi}_1, \dots, \vct{\psi}_{n_{\mtx{\Psi}}}$.}
??? STOP HERE ???

>>>>

All methods considered in this work are based on the following simple idea: Apply an existing randomized trace estimator to $\Trace(\mtx{B}(t))$ with \emph{constant} random vectors, that is, the same randomization is used for each value of the parameter $t$. \textcolor{green}{For example, the Girard-Hutchinson estimator becomes $n_{\mtx{\Psi}}^{-1} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{B}(t) \vct{\psi}_j$ for $n_{\mtx{\Psi}}$ constant Gaussian random vectors $\vct{\psi}_1, \dots, \vct{\psi}_{n_{\mtx{\Psi}}}$.}
??? STOP HERE ???

### 2025-02-17 paper/numresults.tex
adapt nomenclature and fix README

The Krylov-aware stochastic trace estimation \cite{chen-2023-krylovaware-stochastic} is also well suited for approximating the trace of a matrix function $f(\mtx{A}, t)$ which separately depends on a parameter $t$. It also samples two standard Gaussian random matrices $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ and $\mtx{\Psi} \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$. $\mtx{\Omega}$ is used to extract a low-rank approximation to $f(\mtx{A}, t)$ from a block-Krylov subspace of $\mtx{A}$ generated from $r$ iterations of the block-Lanczos method with reorthogonalization and subsequently $q$ without. $\mtx{\Psi}$ is then used to estimate the trace of the approximation residual with $r$ iterations of the Lanczos method.

>>>>

The Krylov-aware stochastic trace estimation \cite{chen-2023-krylovaware-stochastic} is also well suited for approximating the trace of a matrix function $f(\mtx{A}, t)$ which separately depends on a parameter $t$. It also samples two Gaussian random matrices $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ and $\mtx{\Psi} \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$. $\mtx{\Omega}$ is used to extract a low-rank approximation to $f(\mtx{A}, t)$ from a block-Krylov subspace of $\mtx{A}$ generated from $r$ iterations of the block-Lanczos method with reorthogonalization and subsequently $q$ without. $\mtx{\Psi}$ is then used to estimate the trace of the approximation residual with $r$ iterations of the Lanczos method.

### 2025-02-17 paper/spectraldensity.tex
push changes from Overleaf
adapt nomenclature and fix README

The smoothed spectral density $\phi_{\sigma}$ is defined as
\begin{equation}
    \phi_{\sigma}(t) = \frac{1}{n} \sum_{i=1}^{n} g_{\sigma}(t - \lambda_i) = \frac{1}{n} \Trace(g_{\sigma}(t\mtx{I}_n - \mtx{A}))
    \label{equ:smooth-spectral-density}
\end{equation}
for a smoothing kernel $g_{\sigma}$ parametrized by a smoothing parameter $\sigma > 0$, which controls by how much $\phi$ is smoothed. We choose a Gaussian smoothing kernel of width $\sigma > 0$, given by
\begin{equation}
    g_{\sigma}(s) = \frac{1}{\sigma \sqrt{2\pi}} e^{-\sfrac{s^2}{2\sigma^2}}.
    \label{equ:smoothing-kernel}
\end{equation}
The task in \refequ{equ:smooth-spectral-density} is to approximate the trace of the parameter-dependent symmetric positive semi-definite matrix $g_{\sigma}(t\mtx{I}_n - \mtx{A})$. Hence, we can apply the results derived in \refsec{sec:analysis}.

@@

The proof specializes Bernstein's theorem for Chebyshev interpolation to the case of a Gaussian smoothing kernel \refequ{equ:smoothing-kernel}. It loosely follows the proof given in the arXiv preprint of \cite[Theorem 2]{lin-2017-randomized-estimation}, but attains a slightly different dependence on $\sigma$.

@@

    \State Generate standard Gaussian random matrices $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ and $\mtx{\Psi} \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$%\glsfirst{sketching-matrix} $\in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$

>>>>

The smoothed spectral density $\phi_{\sigma}$ is defined as
\begin{equation}
    \phi_{\sigma}(t) = \frac{1}{n} \sum_{i=1}^{n} g_{\sigma}(t - \lambda_i) = \frac{1}{n} \Trace(g_{\sigma}(t\mtx{I}_n - \mtx{A}))
    \label{equ:smooth-spectral-density}
\end{equation}
for a smoothing kernel $g_{\sigma}$ parametrized by a smoothing parameter $\sigma > 0$, which controls by how much $\phi$ is smoothed. We choose a Gaussian smoothing kernel of width $\sigma > 0$, given by
\begin{equation}
    g_{\sigma}(s) = \frac{1}{\sigma \sqrt{2\pi}} e^{-\sfrac{s^2}{2\sigma^2}}.
    \label{equ:smoothing-kernel}
\end{equation}
The task in \refequ{equ:smooth-spectral-density} is to approximate the trace of the parameter-dependent SPSD matrix $g_{\sigma}(t\mtx{I}_n - \mtx{A})$. Hence, we can apply the results derived in \refsec{sec:analysis}.

@@

The proof specializes Bernstein's theorem for Chebyshev interpolation to the case of a Gaussian smoothing kernel \refequ{equ:smoothing-kernel}. It loosely follows the proof given in the arXiv preprint of \cite[Theorem 2]{lin-2017-randomized-estimation}, but attains a slightly different dependence on $\sigma$. \\ \vspace{\parskip}

@@

    \State Generate Gaussian random matrices $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ and $\mtx{\Psi} \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$%\glsfirst{sketching-matrix} $\in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$
